---
title: "StatsFitting5_InvestigationOfSexSplit"
author: "Heywood et al."
output: 
  html_document: 
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This script investigates fitting separate coefficients for each sex for each of the predictors (near-equivalent to having a sex-interaction term, rather than just a sex-offset.) 

Just looking here at the NHM specimens, rather than the much smaller Oxford dataset. 

```{r message = FALSE}
library(tidyverse)
library(brms)
library(posterior)
library(forcats)
library(bayesplot)
library(cowplot)
library(ggrepel)
```

# Creating sex specific datasets

NB this will exclude 'undetermined' specimens!

```{r}
NHM_ForFitting <- read_csv(  '../data/NHM_for_models_monthly.csv')
NHM_males <- NHM_ForFitting %>% filter(Sex == "male")
NHM_females <- NHM_ForFitting %>% filter(Sex == "female")
```

Including only those species we have at least 50 observations of a sex. 

```{r}
TableOfModels_NHM_males <- NHM_males %>%
  count(Sp) %>%
  filter(n > 50)

TableOfModels_NHM_females <- NHM_females %>%
  count(Sp) %>%
  filter(n > 50)
```

# Model Fitting

## Males

```{r eval = FALSE}
Set_males = paste0('Sp.', TableOfModels_NHM_males$Sp, "_males")

CoreModel_males <- NHM_males %>%
  filter(Sp == TableOfModels_NHM_males$Sp[1]) %>%
  brm(log(Right_wing) ~ YearRescale + Temp + Latitude,
      data = ., refresh = 0)

for(i in 1:nrow(TableOfModels_NHM_males)){
  print(Set_males[i])
  NHM_males %>% filter(Sp == TableOfModels_NHM_males$Sp[i]) -> batchDATA
  update(CoreModel_males, newdata = batchDATA, refresh = 0) -> MODEL_FIT
  save(MODEL_FIT,
       file = paste('../NHM_modelfitsM1_SexSplitmales/', Set_males[i]))
}
```

## Females

```{r eval = FALSE}
Set_females = paste0('Sp.', TableOfModels_NHM_females$Sp, "_females")

CoreModel_females <- NHM_females %>%
  filter(Sp == TableOfModels_NHM_females$Sp[1]) %>%
  brm(log(Right_wing) ~ YearRescale + Temp + Latitude,
      data = ., refresh = 0)

for(i in 1:nrow(TableOfModels_NHM_females)){
  print(Set_females[i])
  NHM_females %>%filter(Sp == TableOfModels_NHM_females$Sp[i]) -> batchDATA
  update(CoreModel_females, newdata = batchDATA, refresh = 0) -> MODEL_FIT
  save(MODEL_FIT,
       file = paste('../NHM_modelfitsM1_SexSplitfemales/', Set_females[i]))
}
```

# Posterior summaries

## Male 

```{r eval = FALSE}
map_df(1:nrow(TableOfModels_NHM_males), function(i){
  
  load(file = paste('../NHM_modelfitsM1_SexSplitmales/', Set_males[i]))
  
  MODEL_FIT %>%
    as_draws_array() %>%
    summarise_draws("mean","median","sd","mad","quantile2",
                    ~quantile2(.x, probs =c(0.025,0.167,0.833, 0.975))) -> Fit_df
  
  Fit_df$Sp = TableOfModels_NHM_males$Sp[i]
  Fit_df$Sex = "male"
  
  return(Fit_df)
  
}) -> Posterior_Summaries_NHM_males

write_csv(Posterior_Summaries_NHM_males,
          '../statsresults/NHM_Posterior_Summaries_M1_males.csv')
```

## Female
```{r eval = FALSE}
map_df(1:nrow(TableOfModels_NHM_females), function(i){
  
  load(file = paste('../NHM_modelfitsM1_SexSplitfemales/', Set_females[i]))
  
  MODEL_FIT %>%
    as_draws_array() %>%
    summarise_draws("mean","median","sd","mad","quantile2",
                    ~quantile2(.x, probs = c(0.025,0.167,0.833, 0.975))) -> Fit_df
  
  Fit_df$Sp = TableOfModels_NHM_females$Sp[i]
  Fit_df$Sex = "female"
  
  return(Fit_df)
  
}) -> Posterior_Summaries_NHM_females

write_csv(Posterior_Summaries_NHM_females,
          '../statsresults/NHM_Posterior_Summaries_M1_females.csv')
```

## Joining
```{r eval = FALSE}
bind_rows(Posterior_Summaries_NHM_males,
          Posterior_Summaries_NHM_females) ->
  Posterior_Summaries_NHM_all

write_csv(Posterior_Summaries_NHM_all,
          '../statsresults/NHM_Posterior_Summaries_M1_SexSplitCombined.csv')
```

##  Load posterior summaries
```{r}
PS_NHM_Split <- read_csv('../statsresults/NHM_Posterior_Summaries_M1_SexSplitCombined.csv') %>%
  mutate(Sex = factor(Sex))
```

# Direct Parameter Comparisons

```{r}
full_join(TableOfModels_NHM_females,
          TableOfModels_NHM_males, by = 'Sp') -> JointSplitcounts

## Which Species not with both
JointSplitcounts %>%
  filter( is.na(n.x)|is.na(n.y))

## Which Species not with both
JointSplitcounts %>%
  filter( !(is.na(n.x)|is.na(n.y))) -> SplitSexSpTable

#### Compare (subtract) posteriors for each species:

map_df(1:nrow(SplitSexSpTable), function(i){
  load(file = paste0('../NHM_modelfitsM1_SexSplitmales/Sp.', SplitSexSpTable$Sp[i],  '_males'  ))
  as_draws_array(MODEL_FIT) -> Male_Draws
  
  load(file = paste0('../NHM_modelfitsM1_SexSplitfemales/Sp.', SplitSexSpTable$Sp[i],  '_females'  ))
  as_draws_array(MODEL_FIT) -> Female_Draws
  
  (Male_Draws-Female_Draws) %>%
    summarise_draws("mean","median","sd","mad","quantile2",
                    ~quantile2(.x, probs =c(0.025,0.167,
                                            0.833, 0.975))) -> Draws_Difference_DF
  Draws_Difference_DF$Sp = SplitSexSpTable$Sp[i]
  return(Draws_Difference_DF)
}) -> Posterior_Summaries_NHM_SexDiff


Posterior_Summaries_NHM_SexDiff %>%
  filter(variable %in% c('b_Temp',
                         'b_YearRescale',
                         'b_Latitude')) %>%
  select(variable          ,  mean,Sp,q2.5, q97.5  ) %>% 
  mutate( ConfSignifDiff =   sign(q2.5) ==   sign(q97.5)) -> PosteriorDiffs 

PosteriorDiffs  %>%
  filter( ConfSignifDiff) %>%
  arrange(variable, abs(mean))%>%
  as.data.frame


PosteriorDiffs %>%
  filter( ConfSignifDiff) %>%
  distinct(Sp)
```

# Cross-species impact of predictors: 

## Data loading
```{r}
read_csv('../data/brood_table.csv') %>%
  filter(Include == 'TRUE' )%>%
  select(Sp = `Taxon name` ,Family )-> SpFamily

## Butterfly Red List https://resjournals.onlinelibrary.wiley.com/doi/10.1111/icad.12582
UKRedList <- read_csv('../data/ButterflyRedlistUK.csv')

### Sizes
NHM_ForFitting %>%
  group_by(Sp) %>%
  summarise( MeanRightWing = signif(mean(Right_wing), 4)) -> SpeciesSizeProps

```

## Prepping data
```{r}
PS_NHM_Split %>%
  left_join(SpFamily) %>%
  distinct(Sp, Family ) %>%
  arrange(Sp) -> OrderTable
ORDER_f2 <-pluck(OrderTable,'Sp')

PS_NHM_Split %>%
  mutate(Sp = factor(Sp, levels = ORDER_f2)) %>%
  # mutate( q2.5 = if_else(variable == 'b_Latitude' & Sp == 'Melitaea cinxia', -Inf, q2.5), # Truncate extreme uncertainty (symmetric!)
  #         q97.5 = if_else(variable == 'b_Latitude' & Sp == 'Melitaea cinxia', Inf, q97.5)) %>%
  left_join(SpFamily) %>%
  left_join(UKRedList, by = c('Sp' = 'Species')) %>%
  filter(variable %in% c('b_Temp',
                         'b_YearRescale',
                         'b_Latitude')) %>%
  mutate( Response = case_match(variable,
                                'b_Temp' ~ 'b. Temperature',
                                'b_YearRescale' ~ 'c. Year',
                                'b_Latitude' ~ 'a. Latitude'),
          Sex = factor(Sex)  ) -> DataForMainNHMPlot_SplitSex

### Checking extreme values
DataForMainNHMPlot_SplitSex %>%
  filter(Sp == "Melitaea cinxia", variable == "b_Latitude")
```

```{r message=FALSE, warning=FALSE}
DataForMainNHMPlot_SplitSex %>%
  ggplot(aes( y = fct_rev(Sp),  x = mean, col = Sex )) +
  geom_errorbarh(aes(xmin = q2.5, xmax = q97.5, group = Sex),                            #95% interval not 90
                 linewidth = 0.5, height = 0, col = 'grey',
                 position =  position_dodge(width =0.5)) +
  geom_errorbarh(aes(xmin = q5, xmax = q95, group = Sex ),
                 linewidth = 1, height = 0, col = 'grey',
                 position = position_dodge(width =0.5)) +
  geom_errorbarh(aes(xmin = q16.7 , xmax = q83.3, group = Sex ),
                 linewidth = 2, height = 0, col = 'grey',
                 position = position_dodge(width =0.5)) +
  geom_point(aes(col = Sex), shape = 16,
             position = position_dodge(width =0.5)) +
  geom_vline(xintercept = 0) +
  facet_grid(Family ~ Response,
             scales = 'free',
             space = "free_y") +
  theme_bw() +
  ylab('') +
  xlab('Fitted Coefficient') +
  ggtitle('Species-level responses when split by sex (NHM)') +
  theme( strip.text.y.right = element_text(angle = 0),
         axis.text.y = element_text(face = "italic"),
         strip.background = element_rect(fill = 'grey80', colour = NA),
         legend.position = "bottom" )

ggsave('../figs/NHM_Monthly_MainSpResults_SplitSexSp.png',
       width = 10, height = 9, bg = 'white')

ggsave('../figs/NHM_Monthly_MainSpResults_SplitSexSp.pdf',
       width = 10, height = 9, bg = 'white')
```

# Plotting male and female coefficients against each other
```{r}
SexComparison <- PS_NHM_Split %>%
  filter(variable %in% c("b_Temp", "b_YearRescale", "b_Latitude")) %>%
  select(Sp, Sex, variable, mean, q2.5, q97.5) %>%
  pivot_wider( names_from = Sex,
               values_from = c(mean, q2.5, q97.5),
               names_sep = "_") %>%
  filter(!is.na(mean_male), !is.na(mean_female)) %>%
  mutate(male_confident = sign(q2.5_male) == sign(q97.5_male),
         female_confident = sign(q2.5_female) == sign(q97.5_female),
         difference_type = (sign(mean_male) != sign(mean_female)) &
           (male_confident | female_confident))

SexComparison %>%
  filter(!is.na(mean_male), !is.na(mean_female),Sp != "Melitaea cinxia") %>% # Removing Glanville Fritillary
  ggplot(aes(x = mean_male, y = mean_female)) +
  geom_errorbar(aes(ymin = q2.5_female, ymax = q97.5_female), width = 0) +
  geom_errorbarh(aes(xmin = q2.5_male, xmax = q97.5_male), height = 0) +
  geom_abline(slope = 1, intercept = 0, colour = "grey50") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_point( size = 1, col = 'red') +
  facet_wrap(~variable, scales ='free') +
  theme_bw() +
  theme(aspect.ratio = 1)+
  labs( x = "Male coefficient",
        y = "Female coefficient") 

ggsave('../figs/DirectCompareCoefsBySex.png',
       width = 10, height = 5, bg = 'white')

```

#  Repeating Cross-species Analyses without apparently diverging species. 

```{r}
#Build cross-species dataset
Trends <- read_csv('../statsresults/NHM_Posterior_Summaries_Spatial.csv') %>%
  left_join(UKRedList, by = c('Sp' = 'Species')) %>%
  left_join(SpeciesSizeProps) %>%
  left_join(SpFamily) %>%
  mutate(RedListListed = ifelse(is.na(`GB Red List category`),
                                'Unlisted', 'Listed')) %>%
  filter(variable %in% c('b_Temp', 'b_YearRescale')) %>%
  select(variable, Sp, Family,
         RedListListed,   MeanRightWing,
         mean,sd)

```

## Remove species coefficients for cross-species analyses just for particular analyses.

```{r}
PosteriorDiffs  %>%
  filter( ConfSignifDiff) %>%
  arrange(variable, abs(mean))%>%
  as.data.frame

ToRemove = PosteriorDiffs  %>%
  filter( ConfSignifDiff) %>%
  select( variable,  Sp , ConfSignifDiff)

PS_NHM_WO_SplitSexSp<- Trends %>%
  left_join(ToRemove, by = join_by(variable, Sp)) %>%
  filter( is.na(ConfSignifDiff)  )  # filter out 14 species x variable terms that were confidently different between species

```

## Year coefficient and Size

```{r eval = FALSE}
PS_NHM_WO_SplitSexSp %>%
  filter( variable == 'b_YearRescale' ) %>%
  brm( mean|mi(sd)  ~  MeanRightWing ,
       data = .) -> bYear_model_NHM_Size_WO_SplitSexSp
save(bYear_model_NHM_Size_WO_SplitSexSp, file = '../TopLevelModelFits/bYear_model_NHM_Size_WO_SplitSexSp' )
```

```{r}
load('../TopLevelModelFits/bYear_model_NHM_Size_SPATIAL')  # Called 'MS'
load('../TopLevelModelFits/bYear_model_NHM_Size_WO_SplitSexSp')

#original (all 52 species)
nobs(MS)
posterior_summary(MS)["b_MeanRightWing", ]

#filtered (43 Species )
nobs(bYear_model_NHM_Size_WO_SplitSexSp)
posterior_summary(bYear_model_NHM_Size_WO_SplitSexSp)["b_MeanRightWing", ]
```

## Recreating 4a

```{r}
Orig_sum_4a <-  as.data.frame(posterior_summary(MS))
Filt_sum_4a <-  as.data.frame(posterior_summary(bYear_model_NHM_Size_WO_SplitSexSp))
FillCols = c( 'lightgreen',"blue",'orange',"gold", 'darkgrey', '#800020')
LineCols = c( 'darkgreen',"blue",  'orange', "black", 'black', '#800020')

Trends %>%
  left_join(ToRemove, by = join_by(variable, Sp)) %>%
  mutate( ConfSignifDiff = !is.na(ConfSignifDiff)) %>%
  filter(variable == "b_YearRescale") %>%
  ggplot(aes(x = MeanRightWing, y = mean)) +
  geom_point(aes(col = Family, fill = Family, shape =ConfSignifDiff)) +
 # geom_point(aes(shape =ConfSignifDiff), col = 'black', size = 0.1, da) +
  geom_abline(intercept = Orig_sum_4a$Estimate[1],    ## Original Line
              slope = Orig_sum_4a$Estimate[2], col = 'red')+
  geom_abline(intercept = Filt_sum_4a$Estimate[1],    # Split species removed
              slope = Filt_sum_4a$Estimate[2])+
  scale_shape_manual(values = c( 21,4), name = 'Confident Difference\nBetween Sexes')+
    scale_colour_manual(values = LineCols )+
  scale_fill_manual(values = FillCols )+
  theme_bw() +
  xlab("Mean wing size") +
  ylab("Year coefficient") +
  ggtitle("Sex response test:\nYear Coefficient ~ Size") -> Fig4a_WO_SplitSex

Fig4a_WO_SplitSex
```

## Temp and Red List
```{r eval = FALSE}

PS_NHM_WO_SplitSexSp %>%
  filter( variable == 'b_Temp' ) %>%
  brm( mean|mi(sd)  ~  RedListListed ,
       data = .)  -> bTemp_model_NHM_RedList_WO_SplitSexSp

save(bTemp_model_NHM_RedList_WO_SplitSexSp, file =  '../TopLevelModelFits/bTemp_model_NHM_RedList_WO_SplitSexSp' )
```

```{r}
load('../TopLevelModelFits/bTemp_model_NHM_RedList_SPATIAL') # bTemp_model_NHM_RedList 
load('../TopLevelModelFits/bTemp_model_NHM_RedList_WO_SplitSexSp')

#temperature - original and filtered
#original
nobs(bTemp_model_NHM_RedList)
posterior_summary(bTemp_model_NHM_RedList)["b_RedListListedUnlisted", ]

#filtered
nobs( bTemp_model_NHM_RedList_WO_SplitSexSp)
posterior_summary(bTemp_model_NHM_RedList_WO_SplitSexSp)["b_RedListListedUnlisted", ]
```

## Recreating 4b

```{r}

Trends %>%
  left_join(ToRemove, by = join_by(variable, Sp)) %>%
  filter( variable == 'b_Temp' ) %>%
  mutate( ConfSignifDiff = !is.na(ConfSignifDiff)) -> Fourb_df


Fourb_df%>%
  ggplot(aes(x = RedListListed, y = mean)) +
  geom_boxplot( data = filter(Fourb_df,  !ConfSignifDiff)) +
  geom_point(aes(col = Family, fill = Family, shape =ConfSignifDiff),
             position = position_jitter(width = 0.1)) +
  theme_bw() +
  scale_shape_manual(values = c( 21,4), name = 'Confident Difference\nBetween Sexes')+
  scale_colour_manual(values = LineCols )+
  scale_fill_manual(values = FillCols )+
  xlab("Red List status") +
  ylab("Temperature coefficient") +
  ggtitle("Sex response test:\nTemperature Coefficient ~ Red List") -> Fig4b_WO_SplitSex
Fig4b_WO_SplitSex
```

### Saving Figure

```{r}
Joint4 <- plot_grid(Fig4a_WO_SplitSex+guides(colour = 'none', shape = 'none', fill = 'none'),
                    Fig4b_WO_SplitSex, nrow = 1 )
ggsave(Joint4,  filename = '../figs/Fig4_wosplitsex.png' , height = 5, width = 8)

```


# Session Info

```{r}
sessionInfo()
```
