---
title: "Stats Fitting 2 Across Species NHM Spatial"
author: "Heywood et al."
output: 
  html_document: 
    toc: true
editor_options: 
  chunk_output_type: console
---

Main document for making main text results for the NHM data, comparing across species and fitting further models. Model fits at this level are saved in `TopLevelModelFits/`. 

This document does the analyses for the NHM dataset with the spatial autocorrelation term. See the other Stats Fitting 2 document for the same analyses but using fits without the spatial autocorrelation term.

# Packages and Data Input
```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(brms)
library(posterior)
library(forcats)
library(bayesplot)
library(cowplot)
library(ggrepel)
```

```{r}
## Summaries of posteriors of each species-level model
PS_NHM<- read_csv('../statsresults/NHM_Posterior_Summaries_Spatial.csv')
NHM_ForFitting <- read_csv(  '../data/NHM_for_models_monthly.csv')
```

```{r}
# Specialism 
specialism <- read_csv( '../data/specialism_with_nhm.csv')%>%
  mutate(Label = case_match(Specialism, 
                            'Habitat and foodplant generalist' ~ 'Habitat generalist\n&\nfoodplant generalist',
                            'Habitat and foodplant specialist'~ 'Habitat specialist\n&\nfoodplant specialist',
                            'Habitat generalist and foodplant specialist'~ 'Habitat generalist\n&\nfoodplant specialist',
                            'Habitat specialist and foodplant generalist'~  'Habitat specialist \n&\nfoodplant generalist'))

read_csv('../data/brood_table.csv') %>%
  filter(Include == 'TRUE' )%>%
  select(Sp = `Taxon name` ,Family )-> SpFamily

## Butterfly Red List https://resjournals.onlinelibrary.wiley.com/doi/10.1111/icad.12582
UKRedList <- read_csv('../data/ButterflyRedlistUK.csv')

```


# Prepping data
```{r}
PS_NHM %>%
  left_join(specialism, by =c( 'Sp'='Taxon name')) %>%
  left_join(SpFamily) %>%
  distinct(Specialism, Sp, Family ) %>%
  arrange(Specialism, Sp) -> OrderTable

ORDER_f2 <-pluck(OrderTable,'Sp')

#OrderTable$Family_Cols <- c( 'darkgreen',"blue",  'orange',"gold", 'darkgrey', '#800020')[factor(pluck(OrderTable,'Family'))]
```

NB Glanville fritillary would otherwise overwhelm so truncating uncertainty. 

```{r}
PS_NHM %>%
  mutate(Sp  = factor(Sp, levels = ORDER_f2)) %>% 
  mutate( q5 = if_else(variable== 'b_Latitude' & Sp =='Melitaea cinxia',
                       -Inf, q5)) %>% 
  mutate( q16.7 = if_else(variable== 'b_Latitude' & Sp =='Melitaea cinxia',
                          -Inf, q16.7)) %>%
  left_join(specialism, by =c( 'Sp'='Taxon name')) %>%
  left_join(SpFamily) %>%
  left_join(UKRedList, by =c( 'Sp'='Species') ) %>%
  filter( variable %in% c( 'b_Temp',
                           'b_YearRescale',
                           'b_Sexmale',
                           'b_Latitude') )%>%
  mutate(Response = case_match(variable, 
                               'b_Temp'~ 'b. Temperature',
                               'b_YearRescale'~'c. Year',
                               'b_Sexmale'~'d. Male Offset',
                               'b_Latitude'~'a. Latitude') ) -> DataForMainNHMPlot
```

```{r}
NHM_for_fitting <- read_csv('../data/NHM_for_models_monthly.csv')
NHM_for_fitting %>%
  group_by(Sp)%>%
  summarise(MeanRightWing = signif(mean(Right_wing),4) ,
            SD_RightWing = signif(sd(Right_wing),4),
            CV_RightWing = signif(sd(Right_wing)/mean(Right_wing),4)) -> SpeciesSizeProps

#write_csv(SpeciesSizeProps, '../data/SpeciesSizeProps.csv')### save average butterfly sizes to use for the Oxford data

PS_NHM %>%
  left_join(specialism, by =c( 'Sp'='Taxon name')) %>%
  left_join(UKRedList, by =c( 'Sp'='Species') ) %>%
  left_join(SpeciesSizeProps) %>%
  left_join(SpFamily) %>%
  mutate(RedListListed = ifelse(is.na( `GB Red List category`),
                                'Unlisted',
                                'Listed')) %>% 
  filter( variable %in% c( 'b_Temp',
                           'b_YearRescale') )%>%
  mutate(ConfidentTrend =  sign(q5)== sign(q95),
         TrendSign = ifelse(ConfidentTrend, sign(mean), 0)) %>%
  mutate( Generalist = Specialism == 'Habitat and foodplant generalist') %>%
  select(variable,Sp, Specialism , Family, RedListListed , TrendSign, Generalist,
         mean, sd, MeanRightWing,SD_RightWing ,CV_RightWing) -> Trends

```

# Species-level Results (Main Text Figure)
```{r}
DataForMainNHMPlot %>%
  mutate(Redlist = ifelse( is.na(`GB Red List category`),
                           'Unlisted', 'Listed')) %>%
  ggplot( aes( y =  fct_rev(Sp),
               x = mean))+
  geom_errorbarh(aes( xmin = q5, xmax = q95), linewidth = 1, height = 0, col = 'grey')+
  geom_errorbarh(aes(xmin = q16.7, xmax =  q83.3), linewidth = 2, height = 0, col = 'grey')+
  geom_point(shape = 4)+
  geom_vline(xintercept = 0)+
  facet_grid(Family ~Response, scales = 'free', space = "free_y")+
  ggtitle('With Spatial Random Effect')+
  theme_bw()+
  ylab('')+
  xlab('Fitted Coefficient\n(Log) Wing Size Change, NHM Dataset')+
  theme(strip.text.y.right = element_text(angle = 0),
        axis.text.y = element_text(face =  "italic"),
        strip.background = element_rect(fill = 'grey80', colour = NA))

ggsave('../figs/NHM_Monthly_MainSpResults_byFamily_SPATIAL.png', width = 10, height = 7, bg = 'white')
ggsave('../figs/NHM_Monthly_MainSpResults_byFamily_SPATIAL.pdf', width = 10, height = 7, bg = 'white')
```

## Other Coefficents

Other coefficients of the spatial models

```{r}
PS_NHM %>%
  mutate(Sp  = factor(Sp, levels = ORDER_f2)) %>% 
  left_join(specialism, by =c( 'Sp'='Taxon name')) %>%
  left_join(SpFamily) %>%
  left_join(UKRedList, by =c( 'Sp'='Species') ) %>%
  filter( variable %in% c( 'b_Intercept',
                           'b_Sexundetermined',
                           'sdgp_gpScale_LatScale_Long',
                           'lscale_gpScale_LatScale_Long',
                           'sigma') )%>%
  #### log length scale for clarity
  mutate(   mean   = if_else(variable =='lscale_gpScale_LatScale_Long',log(mean   ),mean   ),
            median = if_else(variable =='lscale_gpScale_LatScale_Long',log(median ),median ),
            sd     = if_else(variable =='lscale_gpScale_LatScale_Long',log(sd     ),sd     ),
            mad    = if_else(variable =='lscale_gpScale_LatScale_Long',log(mad    ),mad    ),    
            q5     = if_else(variable =='lscale_gpScale_LatScale_Long',log(q5     ),q5     ),
            q95    = if_else(variable =='lscale_gpScale_LatScale_Long',log(q95    ),q95    ),
            q16.7  = if_else(variable =='lscale_gpScale_LatScale_Long',log(q16.7  ),q16.7  ),
            q83.3  = if_else(variable =='lscale_gpScale_LatScale_Long',log(q83.3  ),q83.3  )) %>%
  mutate(Response = case_match(variable, 
                               'b_Intercept'~ 'a. Intercept',
                               'b_Sexundetermined'~'b. "Undetermined"\nSex Offset\nfrom Female',
                               'sdgp_gpScale_LatScale_Long'~'c. SD of GP\ncovariance kernel',
                               'lscale_gpScale_LatScale_Long'~'d. Length Scale\nof GP\n(Log-scale)',
                               'sigma'~'e. Residual\nSigma'   ) ) -> DataFor_SI_coefficentPlot

DataFor_SI_coefficentPlot %>%
  ggplot( aes( y =  fct_rev(Sp),
               x = mean))+
  geom_errorbarh(aes( xmin = q5, xmax = q95), linewidth = 1, height = 0, col = 'grey')+
  geom_errorbarh(aes(xmin = q16.7, xmax =  q83.3), linewidth = 2, height = 0, col = 'grey')+
  geom_point(shape = 4)+
  geom_vline(xintercept = 0)+
  facet_grid(Family ~Response, scales = 'free', space = "free_y")+
  theme_bw()+
  ylab('')+
  xlab('Fitted Coefficient\n(Log) Wing Size Change, NHM Dataset')+
  theme(strip.text.y.right = element_text(angle = 0),
        axis.text.y = element_text(face =  "italic"),
        strip.background = element_rect(fill = 'grey80', colour = NA))

ggsave('../figs/SI_NHM_extra_CoefficientPlot.pdf', height = 10, width =10)

```

# Sample Size vs confidence (SI Figure)

```{r}

PS_NHM %>%
  filter( variable %in% c('b_YearRescale' , 'b_Temp'   , 'b_Latitude')) %>%
  left_join(count(NHM_ForFitting, Sp)) %>%
  mutate(ConfidentTrend =  sign(q5)== sign(q95)) %>%
  ggplot(aes( x = n, y = sd))+
  geom_point(aes(col = ConfidentTrend))+
  scale_colour_manual( values = c( 'brown', 'skyblue'),
                       name = 'Confident Trend?')+
  scale_x_sqrt()+
  facet_wrap(~variable, scales = 'free')+
  ylab('SD of posterior of fitted coefficients')+
  xlab('Sample size\n(NB Square-root scale)')+
  theme_bw()+
  theme(legend.position = 'bottom')

ggsave('../figs/Corr_sample_size_trends_NHM_SPATIAL.png', width = 9, height = 4)

```

# Predictors of Year Effect

## Size


```{r eval = FALSE}
### Model Fitting

PS_NHM %>%
  filter( variable == 'b_YearRescale' ) %>%
  brm( mean|mi(sd)  ~  1 ,
       data = .) -> bYear_model_NHM_Average

PS_NHM %>%
  filter( variable == 'b_YearRescale' ) %>%
  left_join(SpeciesSizeProps) %>%
  brm( mean|mi(sd)  ~  MeanRightWing ,
       data = .) -> bYear_model_NHM_Size

### Testing if effect can be explained as well just by family
PS_NHM %>%
  filter( variable == 'b_YearRescale' ) %>%
  left_join(SpeciesSizeProps) %>%
  left_join(SpFamily) %>%
  brm( mean|mi(sd)  ~  Family ,
       data = .) ->  bYear_model_NHM_Family

MS <- add_criterion(bYear_model_NHM_Size, criterion = 'loo')
MF <- add_criterion(bYear_model_NHM_Family, criterion = 'loo')


save(bYear_model_NHM_Average,file= '../TopLevelModelFits/bYear_model_NHM_Average_SPATIAL')
save(MS, file='../TopLevelModelFits/bYear_model_NHM_Size_SPATIAL')
save(MF, file='../TopLevelModelFits/bYear_model_NHM_Family_SPATIAL')
```

```{r warning=FALSE}
load('../TopLevelModelFits/bYear_model_NHM_Average_SPATIAL')
load('../TopLevelModelFits/bYear_model_NHM_Size_SPATIAL')
load('../TopLevelModelFits/bYear_model_NHM_Family_SPATIAL')
```

###  What is average effect of 'Year_rescale'

```{r}
bYear_model_NHM_Average %>% posterior_summary()
```

### Effect of Size on Year Coefficient

```{r}
MS %>% posterior_summary() %>% as.data.frame()-> SizeEffectSummary
SizeEffectSummary

mcmc_hist(MS,
          pars = 'b_MeanRightWing')+
  xlab('Slope of Average Size vs bYear Coefficient,\naccounting for uncertainty')+
  theme_bw()

```

### How does Size compare as a predictor to family?

```{r}
loo_compare(MS, MF)  # Can't differentiate 
```

## Other non-significant predictors (Conservation status + specialism)

```{r eval = FALSE}
Trends %>%
  filter( variable == 'b_YearRescale' ) %>%
  brm( mean|mi(sd)  ~   Generalist     ,
       data = .) -> bYear_model_NHM_Specialism

Trends %>%
  filter( variable == 'b_YearRescale' ) %>%
  brm( mean|mi(sd)  ~   RedListListed      ,
       data = .) -> bYear_model_NHM_RedList

save(bYear_model_NHM_Specialism, file = '../TopLevelModelFits/bYear_model_NHM_Specialism_SPATIAL')
save(bYear_model_NHM_RedList, file = '../TopLevelModelFits/bYear_model_NHM_RedList_SPATIAL')
```

```{r warning=FALSE}
load('../TopLevelModelFits/bYear_model_NHM_Specialism_SPATIAL')
load('../TopLevelModelFits/bYear_model_NHM_RedList_SPATIAL')

bYear_model_NHM_Specialism %>% posterior_summary()
bYear_model_NHM_RedList    %>% posterior_summary()

```

# Predictors of Effect of Late-Larval Temperature:

## Models for Redlist size (/family)  + specialism

```{R eval = FALSE}
PS_NHM %>%
  filter( variable == 'b_Temp' ) %>%
  left_join(UKRedList, by =c( 'Sp'='Species') ) %>%
  mutate(RedListListed = ifelse(is.na( `GB Red List category`),
                                'Unlisted', 'Listed')) %>%
  brm( mean|mi(sd)  ~  RedListListed ,
       data = .) -> bTemp_model_NHM_RedList

Trends %>%
  filter( variable == 'b_Temp' ) %>%
  brm( mean|mi(sd)  ~   Generalist     ,
       data = .) -> bTemp_model_NHM_Specialism

Trends %>%
  filter( variable == 'b_Temp' ) %>%
  brm( mean|mi(sd)  ~   MeanRightWing      ,
       data = .) -> bTemp_model_NHM_Size


save(bTemp_model_NHM_RedList, file = '../TopLevelModelFits/bTemp_model_NHM_RedList_SPATIAL')
save(bTemp_model_NHM_Specialism, file = '../TopLevelModelFits/bTemp_model_NHM_Specialism_SPATIAL')
save(bTemp_model_NHM_Size, file = '../TopLevelModelFits/bTemp_model_NHM_Size_SPATIAL')
```

```{r warning=FALSE}
load('../TopLevelModelFits/bTemp_model_NHM_RedList_SPATIAL')
load('../TopLevelModelFits/bTemp_model_NHM_Specialism_SPATIAL')
load('../TopLevelModelFits/bTemp_model_NHM_Size_SPATIAL')

bTemp_model_NHM_RedList %>% posterior_summary()
bTemp_model_NHM_Specialism %>% posterior_summary()
bTemp_model_NHM_Size %>% posterior_summary()

```

```{r}
Trends %>% 
  group_by(variable, RedListListed)%>%
  count( TrendSign) %>%
  pivot_wider(names_from = RedListListed,
              id_cols =c(variable,TrendSign),
              values_from = n ) %>%
  t %>%
  knitr::kable()

#### Chi-Squared Red list and temp effect:
chisq.test(matrix(c(9,4,
                    13,13,
                    4,9),3,2, byrow = TRUE))
```

# Generating Combined Figure for paper

```{r}
FillCols = c( 'lightgreen',"blue",'orange',"gold", 'darkgrey', '#800020')
LineCols = c( 'darkgreen',"blue",  'orange', "black", 'black', '#800020')

Trends %>%
  filter( variable == 'b_YearRescale' ) %>%
  ggplot(aes( x = MeanRightWing, y= mean))+
  geom_abline(intercept = SizeEffectSummary$Estimate[1],
              slope = SizeEffectSummary$Estimate[2])+
  geom_point(aes(col = Family, fill = Family), shape = 21)+
  theme_bw()+
  scale_fill_manual(values = FillCols )+
  scale_colour_manual(values = LineCols )+
  theme(legend.position = 'bottom')+
  xlab('Butterfly Average Size\n(Mean Right Wing, mm)')+
  ylab('"Year" Coefficient') ->YearEffect_Size_Plot_nolabels

YearEffect_Size_Plot_nolabels

Trends %>%
  filter( variable == 'b_Temp' ) %>%
  ggplot(aes( x = RedListListed, y= mean))+
  geom_boxplot(colour = 'grey70', fill = 'grey90', linewidth = 2,
               outlier.shape = NA)+
  geom_hline( yintercept = 0, linetype = 'dashed')+
  geom_point(aes(col = Family,fill = Family), shape = 21,
             position = position_jitter(width = 0.15, seed = 1 ))+
  scale_colour_manual(values = LineCols )+
  scale_fill_manual(values = FillCols )+
  theme_bw()+
  theme(legend.position = 'bottom')+
  xlab('UK Red List\nStatus')+
  ylab('"Temperature" Coefficient')  -> TempEffect_RedList_Plot_nolabels

TempEffect_RedList_Plot_nolabels

plot_grid(plot_grid( YearEffect_Size_Plot_nolabels +
                       guides( col = 'none', fill = 'none'),
                     TempEffect_RedList_Plot_nolabels+
                       guides( col = 'none', fill = 'none'),
                     labels = c('a)', 'b)') ),
          get_legend(TempEffect_RedList_Plot_nolabels ),
          ncol=1 , rel_heights = c(5,1))

ggsave('../figs/PredictorsFigure_Both_nolabels_SPATIAL.png',
       width = 8, height = 5, dpi = 300, bg = 'white')

```

## Repeating With labels (SI Figure)
```{r}

YearEffect_Size_Plot_WITHlabels <- YearEffect_Size_Plot_nolabels+
  geom_errorbar(aes( ymax = mean+(2*sd),  ymin = mean-(2*sd)), col = 'grey70')+
  scale_x_continuous(limits = c(10, NA))+
  geom_text_repel(aes(label = Sp), size = 2 ,hjust=-0.15,
                  box.padding = unit(0.2, "lines"),
                  direction = 'y', seed=1 , max.overlaps = Inf)
   
YearEffect_Size_Plot_WITHlabels

ggsave(YearEffect_Size_Plot_WITHlabels,
       filename = '../figs/PredictorsFigure_4_withlabels_SPATIAL.png',
       width = 8, height = 12, dpi = 300, bg = 'white')
```

# Session Info

```{r}
sessionInfo()
```