---
title: "StatsFitting1_SpeciesLevel"
author: "Heywood et al."
output: 
  html_document: 
    toc: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This script fits and saves `brms` models predicting size for each species separately.

The core model is: log(Size)  ~ YearRescale + Temp  + Latitude + Sex. Alternative models that include a 'spatial random effect' by fitting a Gaussian Point Processes to capture additional autocorrelation in sizes explainable by the proximity of locations are also tested. 

Towards the end of this the improvement in fit that the spatial-autocorrelation term bring is assessed and the fitted coefficients with and without compared . 

Models testing fitting distinct responses to the predictors for each sex afre fit in a seperate document (`StatsFitting5`)

`brms` model objects are saved in folders, and summaries of the posteriors are saved as .csvs for further use. 

```{r message = FALSE}
library(tidyverse)
library(brms)
library(posterior)
```

```{r}
NHM_ForFitting <- read_csv(  '../data/NHM_for_models_monthly.csv')
OUM_ForFitting <- read_csv(  '../data/OUMNH_for_models_monthly.csv')
```

# NHM Fit

## Identify species with enough data
```{r}
Batches<- count(NHM_ForFitting, Sp)%>% arrange( n)

Batches$n %>% log10 %>% plot; abline(h = 2) # almost perfect log-normal plot
sum(Batches$n >100) # 52 species 

Batches %>%
  filter( n >100 ) -> TableOfModels_NHM

median(TableOfModels_NHM$n)

Set2 = paste0('Sp.', TableOfModels_NHM$Sp)
```

### Spatial aggregation and rescaling of lat longs. 

In the NHM dataset, lats and longs are given to implausibly precise 4dp precision (approx 10m), presumably due to some transformation along the way. Because fitting time scales with the cube of the number of distinct locations, it is worth just checking if moderate spatial aggregation of very nearby points impacts the number of distinct collection locations. 

```{r}
# How many unique locations in largest dataset?
full_join(NHM_ForFitting %>%
            distinct( Sp, Latitude, Longitude ) %>%
            count(Sp, sort = TRUE, name = 'Original4dpResCount'),
          NHM_ForFitting %>% ## rounding to nearest km or 10km doesn't affect these counts much.
            mutate( LatitudeRound = round( Latitude, digits = 1), #  Round to nearest ~10km
                    LongitudeRound = round( Longitude, digits = 1)) %>%
            distinct( Sp, LatitudeRound, LongitudeRound ) %>%
            count(Sp, sort = TRUE, name = 'Coarse1dpResCount'))
```

Since coarsening resolution only makes a moderate reduction in sites, will just fit on original resolution.

Scaling lat and longs across all the species for slightly more consistent values. NB that `gp()` also includes a further (within species) scaling (to make max distance = 1), so the impact is relatively minor. 

```{r}
NHM_ForFitting %>%
  mutate( Scale_Lat = scale( Latitude)[,1],
          Scale_Long= scale(Longitude)[,1]) -> NHM_ForFitting
```

## Fitting Models With Spatial Term
```{r eval = FALSE}
### Define and compile model for first species
CoreModelSP <-   NHM_ForFitting %>%
    filter(Sp == TableOfModels_NHM$Sp[1]) %>%
    brm( log(Right_wing)  ~ YearRescale + Temp  + Latitude + Sex +
           gp(Scale_Lat, Scale_Long,
              cov = "exp_quad", #  Default exponentiated-quadratic kernel
              gr = TRUE),  ## Group observations at the same location
         data = ., 
         refresh = 0)

## Refit model to each set of data, saving each time
for( i in 1:nrow( TableOfModels_NHM)){
  print(Set2[i])
  print( Sys.time())
  NHM_ForFitting %>%
    filter(Sp == TableOfModels_NHM$Sp[i]) -> batchDATA
  update( CoreModelSP, newdata = batchDATA, refresh = 0) -> MODEL_FIT
  save(MODEL_FIT, file = paste0('../NHM_modelfitsSpatial/',Set2[i]  ))
}

```

### Checking model fit

```{R}
MaxRHats_NHM <- map_df(1:nrow( TableOfModels_NHM), function(i){
  load(file = paste0('../NHM_modelfitsSpatial/',Set2[i]  )) 
  return( data.frame(Sp =Set2[i],  MaxRhat =  max(rhat(MODEL_FIT)) ) )
})

arrange(MaxRHats_NHM, desc(MaxRhat)) # Glanville Fritillary not converged well
```

### Fixing Glanville Fritillary Convergence issue

From inspection, clearly problem of GPP sd parameter getting stuck far too high on one of the chains. Dropping dodgy first chain (still leaving three that mixed well) is a neater solution than fiddling with priors, or just rerunning the whole analysis and hoping.
 
```{r eval = FALSE}
 load(file='../NHM_modelfitsSpatial/ Sp.Melitaea cinxia')

 MODEL_FIT
 mcmc_plot(MODEL_FIT, type= 'trace' ) 

new_fit_wo_stuckchain<- MODEL_FIT

  sim = new_fit_wo_stuckchain$fit@sim
  sim$samples <- sim$samples[-1]
  sim$chains = sim$chains - 1
  sim$warmup2 = sim$warmup2[-1]
  new_fit_wo_stuckchain$fit@sim = sim

  ## Good fit now:
 new_fit_wo_stuckchain
  mcmc_plot(new_fit_wo_stuckchain, type= 'trace' ) 

  MODEL_FIT<- new_fit_wo_stuckchain
save(MODEL_FIT, file='../NHM_modelfitsSpatial/ Sp.Melitaea cinxia')
```

### Collating model outputs into a df

```{r eval = FALSE}
map_df(1:nrow( TableOfModels_NHM), function(i){
  load(file = paste0('../NHM_modelfitsSpatial/',Set2[i]  )) 
  MODEL_FIT %>%
    as_draws_array %>%
    summarise_draws( "mean","median","sd","mad","quantile2",
                    ~quantile2(.x, probs = c(0.167, 0.833))) -> Fit_df
  Fit_df$Sp = TableOfModels_NHM$Sp[i]
  return(Fit_df)
})-> Posterior_Summaries_NHM_Spatial

write_csv(Posterior_Summaries_NHM_Spatial, '../statsresults/NHM_Posterior_Summaries_Spatial.csv')
```

## Repeating Fitting Models without spatial term, for model testing. 

```{r eval = FALSE}
### Define and compile model
CoreModel2 <-   NHM_ForFitting %>%
    filter(Sp == TableOfModels_NHM$Sp[1]) %>%
    brm( log(Right_wing)  ~ YearRescale + Temp  + Latitude + Sex,
         data = ., refresh = 0)

## Refit model to each set of data, saving each time
for( i in 1:nrow( TableOfModels_NHM)){
  print(Set2[i])
  NHM_ForFitting %>%
    filter(Sp == TableOfModels_NHM$Sp[i]) -> batchDATA
  update( CoreModel2, newdata = batchDATA, refresh = 0) -> MODEL_FIT
  save(MODEL_FIT, file = paste0('../NHM_modelfitsM1/',Set2[i]  ))
}
```

### Collating model outputs into a df

```{r eval = FALSE}
map_df(1:nrow( TableOfModels_NHM), function(i){
  load(file = paste0('../NHM_modelfitsM1/',Set2[i]  )) 
  MODEL_FIT %>%
    as_draws_array %>%
    summarise_draws( "mean","median","sd","mad","quantile2",
                    ~quantile2(.x, probs = c(0.167, 0.833))) -> Fit_df
  Fit_df$Sp = TableOfModels_NHM$Sp[i]
  return(Fit_df)
})-> Posterior_Summaries_NHM2
write_csv(Posterior_Summaries_NHM2, '../statsresults/NHM_Posterior_Summaries_M1.csv')
```

## Correlations between predictors
For each species calculate correlation between predictors
```{r}
map_df(TableOfModels_NHM$Sp ,function(sp_i){
  
  NHM_ForFitting%>%
    filter(Sp == sp_i) %>%
    select(Latitude ,Year,Temp) %>%
    cor ->x
  return(data.frame( Species = sp_i,
                       YvL = x[2,1],
                     YvT = x[2,3],
                     LvT = x[1,3]))
})-> SpeciesLvlCorrs

par(mfrow= c(1,3))
SpeciesLvlCorrs$YvL %>% hist(breaks=20, main ='Year v Latitude')
abline(v=0, col = 'red'); abline(v=mean(SpeciesLvlCorrs$YvL), col = 'blue')
SpeciesLvlCorrs$YvT %>% hist(breaks=20, main ='Year\nv\nTemperature', xlab = 'Correlation Coefficient')
abline(v=0, col = 'red'); abline(v=mean(SpeciesLvlCorrs$YvT), col = 'blue')
SpeciesLvlCorrs$LvT %>% hist(breaks=20, main ='Temperature v Latitude')
abline(v=0, col = 'red'); abline(v=mean(SpeciesLvlCorrs$LvT), col = 'blue')

png(file ='../figs/SI_corr_inTempLatYear.png',
    units ='in', width = 5, height = 3, res = 400)
par(mfrow= c(1,3))
SpeciesLvlCorrs$YvL %>% hist(breaks=20, main ='Year\nv\nLatitude')
abline(v=0, col = 'red'); abline(v=mean(SpeciesLvlCorrs$YvL), col = 'blue')
SpeciesLvlCorrs$YvT %>% hist(breaks=20, main ='Year\nv\nTemperature', xlab = 'Correlation Coefficient')
abline(v=0, col = 'red'); abline(v=mean(SpeciesLvlCorrs$YvT), col = 'blue')
SpeciesLvlCorrs$LvT %>% hist(breaks=20, main ='Temperature\nv\nLatitude')
abline(v=0, col = 'red'); abline(v=mean(SpeciesLvlCorrs$LvT), col = 'blue')
dev.off()

```

# OUMNH Model fitting

## Identify species with enough data

```{r}
## Cycle through each species to fit models 
Batches<- count(OUM_ForFitting,Sp)%>%
  arrange( desc(n))
Batches$n  %>% plot; abline(h = 40) # almost perfect log-normal plot

Batches %>%
  filter( n >1) -> TableOfModels_OUMNH ## No longer include a threshold

TableOfModels_OUMNH$n %>% median

Set2 = paste0('Sp.', TableOfModels_OUMNH$Sp)

```

## Model with Spatial Autocorrelation Term

NB some species are not differentiated at all by sex, so for those species need to fit a model without a sex-intercept. 

```{r eval = FALSE}

OUM_ForFitting %>%
  mutate( Scale_Lat = scale( Latitude)[,1],
          Scale_Long= scale(Longitude)[,1]) -> OUM_ForFitting

CoreModel_OU_splitSex <-   OUM_ForFitting %>%
  filter(Sp == TableOfModels_OUMNH$Sp[4]) %>%
  brm( log(Wing_Length)   ~ YearRescale + Temp  + Latitude + Sex +
           gp(Scale_Lat ,  Scale_Long ,
              cov = "exp_quad", #  Default exponentiated-quadratic kernel
              gr = TRUE),  ## Group observations at the same location
       data = ., refresh = 0)

CoreModel_OU_jointSex <-   OUM_ForFitting %>%
  filter(Sp == TableOfModels_OUMNH$Sp[2]) %>%
  brm( log(Wing_Length)   ~ YearRescale + Temp  + Latitude +
           gp(Scale_Lat , Scale_Long ,
              cov = "exp_quad", #  Default exponentiated-quadratic kernel
              gr = TRUE),  ## Group observations at the same location,
       data = ., refresh = 0)

undetermined_sp<- distinct(OUM_ForFitting, Sp, Sex) %>% filter( Sex == 'undetermined') %>% pluck('Sp')

## Refit model to each set of data, saving each time
for( i in 1:nrow( TableOfModels_OUMNH)){
  print(Set2[i])
  if(TableOfModels_OUMNH$Sp[i] %in% undetermined_sp ){
    OUM_ForFitting %>%
      filter(Sp == TableOfModels_OUMNH$Sp[i]) -> batchDATA
    update( CoreModel_OU_jointSex, newdata = batchDATA, refresh = 0,
            iter = 10000 ) -> MODEL_FIT
    
  }else{
    OUM_ForFitting %>%
      filter(Sp == TableOfModels_OUMNH$Sp[i]) -> batchDATA
    update( CoreModel_OU_splitSex, newdata = batchDATA, refresh = 0,
            iter = 10000 ) -> MODEL_FIT
  }
  save(MODEL_FIT, file = paste0('../OUM_modelfitsSpatial/',Set2[i]  ))
}
```

### Checking model fit

```{R}
MaxRHats <- map_df(1:nrow( TableOfModels_OUMNH), function(i){
  load(file = paste0('../OUM_modelfitsSpatial/',Set2[i]  )) 
  return( data.frame(Sp =Set2[i],  MaxRhat =  max(rhat(MODEL_FIT)) ) )
})

TableOfModels_OUMNH %>%
  select(n) %>%
  add_column('Sp' = Set2) %>%
  left_join(MaxRHats ) %>%
arrange(desc(MaxRhat)) -> N_v_MaxRhat

N_v_MaxRhat

## inspecting worse case (only 25 specimens...)
load('../OUM_modelfitsSpatial/Sp.Celastrina argiolus')
mcmc_plot(MODEL_FIT, type= 'trace' )  ## problem of GPP sd getting stuck far too high. Will just leave as end up not finding spatial term helpful and just using the direct regression.
```

### Collating model outputs into a df

```{r eval = FALSE}
map_df(1:nrow( TableOfModels_OUMNH), function(i){
  load(file = paste0('../OUM_modelfitsM1/',Set2[i]  )) 
  MODEL_FIT %>%
    as_draws_array %>%
    summarise_draws("mean","median","sd","mad","quantile2",
                    ~quantile2(.x, probs = c(0.167, 0.833))) -> Fit_df
  Fit_df$Sp = TableOfModels_OUMNH$Sp[i]
  return(Fit_df)
})-> Posterior_Summaries_OUM_Spatial

write_csv(Posterior_Summaries_OUM_Spatial, '../statsresults/OUM_Posterior_Summaries_Spatial.csv')

```

## Without Spatial Autcorrelation

```{r eval = FALSE}
CoreModel_OU_splitSex <-   OUM_ForFitting %>%
  filter(Sp == TableOfModels_OUMNH$Sp[4]) %>%
  brm( log(Wing_Length)   ~ YearRescale + Temp  + Latitude + Sex,
       data = ., refresh = 0)

CoreModel_OU_jointSex <-   OUM_ForFitting %>%
  filter(Sp == TableOfModels_OUMNH$Sp[2]) %>%
  brm( log(Wing_Length)   ~ YearRescale + Temp  + Latitude ,
       data = ., refresh = 0)

undetermined_sp<- distinct(OUM_ForFitting, Sp, Sex) %>% filter( Sex == 'undetermined') %>% pluck('Sp')

## Refit model to each set of data, saving each time
for( i in 1:nrow( TableOfModels_OUMNH)){
  print(Set2[i])
  if(TableOfModels_OUMNH$Sp[i] %in% undetermined_sp ){
    OUM_ForFitting %>%
      filter(Sp == TableOfModels_OUMNH$Sp[i]) -> batchDATA
    update( CoreModel_OU_jointSex, newdata = batchDATA, refresh = 0) -> MODEL_FIT
    
  }else{
    OUM_ForFitting %>%
      filter(Sp == TableOfModels_OUMNH$Sp[i]) -> batchDATA
    update( CoreModel_OU_splitSex, newdata = batchDATA, refresh = 0) -> MODEL_FIT
  }
  save(MODEL_FIT, file = paste0('../OUM_modelfitsM1/',Set2[i]  ))
}
```

### Collating model outputs into a df

```{r eval = FALSE}
map_df(1:nrow( TableOfModels_OUMNH), function(i){
  load(file = paste0('../OUM_modelfitsM1/',Set2[i]  )) 
  MODEL_FIT %>%
    as_draws_array %>%
    summarise_draws("mean","median","sd","mad","quantile2",
                    ~quantile2(.x, probs = c(0.167, 0.833))) -> Fit_df
  Fit_df$Sp = TableOfModels_OUMNH$Sp[i]
  return(Fit_df)
})-> Posterior_Summaries_OUM2

write_csv(Posterior_Summaries_OUM2, '../statsresults/OUM_Posterior_Summaries_M1.csv')

```

# Testing Influence of Spatial Autocorrelation term. 

## NHM Model Compare

Comparing models with ELPD. Need to add criterion to the model fits, so takes about 10 minutes to run. (.csv outputs saved for reuse. )

```{r eval = FALSE}
count(NHM_ForFitting,Sp) %>%
  filter( n >100 ) -> TableOfModels_NHM
Set2 = paste0('Sp.', TableOfModels_NHM$Sp)

map_df(1:52, function(i){
  
  load(file = paste0('../NHM_modelfitsSpatial/',Set2[i]  )) 
  SPATIAL_MF <- MODEL_FIT
  
  load(file = paste0('../NHM_modelfitsM1/',Set2[i]  )) 
  ORIG_MF <- MODEL_FIT
  
  SPATIAL_MF<-  add_criterion(SPATIAL_MF, "loo")
  ORIG_MF<-  add_criterion(ORIG_MF,    "loo")
  
  XX<- rownames_to_column(as.data.frame(loo_compare(SPATIAL_MF, ORIG_MF)))
  XX$Sp = TableOfModels_NHM$Sp[i]
  print(TableOfModels_NHM$Sp[i])
  return(XX)
})-> LOO_CompareDF
write_csv(LOO_CompareDF, '../statsresults/SpatialLOOCompares.csv')
```

```{r}
LOO_CompareDF<-read_csv('../statsresults/SpatialLOOCompares.csv')

LOO_CompareDF%>% 
  select(rowname,elpd_diff,se_diff, Sp) %>%
  filter( elpd_diff <0) %>%
  mutate(ConfidentDiff = elpd_diff < -4 & (-elpd_diff > (2*se_diff))) %>%
  arrange(ConfidentDiff) %>%
  rename('Worse Model' = rowname ) %>%
  knitr::kable()
```
    
### Comparing Coefficients 

```{R}
Posterior_Summaries_NHM_SPATIAL<-read_csv('../statsresults/NHM_Posterior_Summaries_Spatial.csv')
Posterior_Summaries_NHM_SPATIAL$Fit <- 'Spatial'

Posterior_Summaries_NHM_ORIG <- read_csv('../statsresults/NHM_Posterior_Summaries_M1.csv')
Posterior_Summaries_NHM_ORIG$Fit <- 'Orig_NoSpatial'

Both_Posteriors <- bind_rows(Posterior_Summaries_NHM_SPATIAL, Posterior_Summaries_NHM_ORIG)

Both_Posteriors %>%
  mutate( q5 = if_else(variable== 'b_Latitude' & Sp =='Melitaea cinxia',
                       -Inf, q5)) %>% 
  mutate( q16.7 = if_else(variable== 'b_Latitude' & Sp =='Melitaea cinxia',
                          -Inf, q16.7)) %>%
  filter( variable %in% c( 'b_Temp',
                           'b_YearRescale',
                           'b_Sexmale',
                           'b_Latitude') )%>%
  mutate(Response = case_match(variable, 
                               'b_Temp'~ 'b. Temperature',
                               'b_YearRescale'~'c. Year',
                               'b_Sexmale'~'d. Male Offset',
                               'b_Latitude'~'a. Latitude') ) -> DataForMainNHMPlot
```


```{r fig.height=10, fig.width=10}
DataForMainNHMPlot%>%
  ggplot( aes( y =  fct_rev(Sp),
               x = mean, col = Fit, group= Fit))+
  geom_errorbarh(aes( xmin = q5, xmax = q95),
                 linewidth = 0.5, height = 0,
                 position =  position_dodge(width =0.5))+
  geom_errorbarh(aes(xmin = q16.7, xmax =  q83.3),
                 linewidth = 1, height = 0,
                 position =  position_dodge(width =0.5))+
  geom_point(shape = 4,
                 position =  position_dodge(width =0.5), col = 'black')+
  geom_vline(xintercept = 0)+
  facet_grid(~Response, scales = 'free', space = "free_y")+
    scale_color_manual(labels = c('Without'  , 'With'),
                       values= c('gold2', 'lightblue'), name = 'Spatial Autocorrelation Fit:')+
  ggtitle('')+
  theme_bw()+
  ylab('')+
  xlab('Fitted Coefficient')+
  theme(strip.text.y.right = element_text(angle = 0),
        legend.position = 'bottom',
        axis.text.y = element_text(face =  "italic"),
        strip.background = element_rect(fill = 'grey80', colour = NA))

 ggsave('../figs/Compare_Spatial_NonSpatial.pdf', height = 10, width = 10)
 ggsave('../figs/Compare_Spatial_NonSpatial.png', height = 10, width = 10)

```

## OUM Spatial Autocorrelation Model comparisons. 

NB much smaller datasets, but still takes a few minutes to calculate the LOO critera. 
Quite a lot of the species have 1-2 very outlying observations that trigger concerns based on expectation of Pareto distributions. 

```{r eval = FALSE}
count(OUM_ForFitting,Sp)%>%
  filter( n >1) -> TableOfModels_OUMNH ## No threshold
Set2 = paste0('Sp.', TableOfModels_OUMNH$Sp)

map_df(1:28, function(i){
  
  load(file = paste0('../OUM_modelfitsSpatial/',Set2[i]  )) 
  SPATIAL_MF <- MODEL_FIT
  
  load(file = paste0('../OUM_modelfitsM1/',Set2[i]  )) 
  ORIG_MF <- MODEL_FIT
  
  SPATIAL_MF<-  add_criterion(SPATIAL_MF, "loo")
  ORIG_MF<-  add_criterion(ORIG_MF,    "loo")
  
  XX<- rownames_to_column(as.data.frame(loo_compare(SPATIAL_MF, ORIG_MF)))
  XX$Sp = TableOfModels_OUMNH$Sp[i]
  print(TableOfModels_OUMNH$Sp[i])
  return(XX)
})-> LOO_CompareDF_OUM
write_csv(LOO_CompareDF_OUM, '../statsresults/SpatialLOOCompares_OUM.csv')

```
  
Only one of the species could demonstrate a confident case to include the random effect.   
  
Also note that a few of the species the convergence was a bit ropey with the spatial autocorrelation term.   
  
```{r}
LOO_CompareDF_OUM<-read_csv('../statsresults/SpatialLOOCompares_OUM.csv')

LOO_CompareDF_OUM%>% 
  select(rowname,elpd_diff,se_diff, Sp) %>%
  filter( elpd_diff <0) %>%
  mutate(ConfidentDiff = elpd_diff < -4 & (-elpd_diff > (2*se_diff))) %>%
  arrange(ConfidentDiff) %>%
  rename('Worse Model' = rowname ) %>%
  knitr::kable()
  
```

# Session Info

```{R}
sessionInfo()

```